---
title: "How to update database with new article data"
date: "Latest update: June 2023"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{pathogen_database_update}
  %\usepackage[utf8]{inputenc}
---

As new articles on WHO priority pathogens are published we provide the ability for people to update the database with new data. New data will be represented as new rows in the four data sets (articles, outbreaks, parameters and models) and any new article will have a new row in the articles dataset and potentially multiple rows in any of the others. We would suggest the below process to add new data, however the only requirement is a submission of a pull request with the updated csv files together with details of the peer-reviewed paper which is being added. The paper should conform with the inclusion and exclusion criteria of the underlying systematic review.

## Step 1 -- git clone repo

Please clone the git repository ``git clone https://github.com/mrc-ide/epireview.git`` and create a new branch to add data for the paper(s)

## Step 2 -- add article details

Please provide all available data for the article. For the quality assessment we refer to the wiki on the [epireview github repo](https://github.com/mrc-ide/epireview/wiki/Quality-assessment) which has details on how to assess quality for paper (the score should be 0, 1 or NA if not applicable).

```{r add article, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# load packages and source code
library(tidyverse)
library(validate)
source('../R/create_new_article_entry.R')
source('../R/append_new_entry_to_table.R')
source('../R/load_data.R')

new_article <- create_new_article_entry('marburg',new_article = c( list( "first_author_first_name" = as.character("Joe")),
                                                    list( "first_author_surname"    = as.character("Blocks")),
                                                    list( "article_title"           = as.character("hello")),
                                                    list( "doi"                     = as.character(NA)),
                                                    list( "journal"                 = as.character("ABC")),
                                                    list( "year_publication"        = as.integer(2000)),
                                                    list( "volume"                  = as.integer(NA)),
                                                    list( "issue"                   = as.integer(NA)),
                                                    list( "page_first"              = as.integer(NA)),
                                                    list( "page_last"               = as.integer(NA)),
                                                    list( "paper_copy_only"         = as.logical(NA)),
                                                    list( "notes"                   = as.character(NA)),
                                                    list( "qa_m1"                   = as.integer(1)),
                                                    list( "qa_m2"                   = as.integer(0)),
                                                    list( "qa_a3"                   = as.integer(NA)),
                                                    list( "qa_a4"                   = as.integer(1)),
                                                    list( "qa_d5"                   = as.integer(0)),
                                                    list( "qa_d6"                   = as.integer(NA)),
                                                    list( "qa_d7"                   = as.integer(1))),
                                        vignette_prepend = "")

```

This creates a new entry for this article. We enforce that first author initial or first name, surname, article title, journal and year of publication needs to be provided. Please take note of the `article_id` and `covidence_id` as you will need these to link any parameters, outbreaks or models to this paper. This can be appended to the articles dataset by running the following command (which included validations for the data provided):

```{r eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
new_article_dataset <- append_new_entry_to_table(pathogen = 'marburg', table_type = 'article', new_row = new_article,
                          validate = TRUE, write_table = TRUE, vignette_prepend = "" )
```

## Step 3 -- add model details

If the paper does not include a model you can skip this section. For this table we have a set number of field options which can be accepted. These field options can be viewed in the `Database field options` vignette. The validations in the function will assert that valid options have been used.

```{r add model, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# load packages and source code
library(tidyverse)
library(validate)
source('../R/create_new_model_entry.R')

new_model <- create_new_model_entry('marburg',c( list( "article_id"           = as.integer(1)),
                                    list( "model_type"           = as.character("Compartmental")),
                                    list( "compartmental_type"   = as.character("SEIR,SIR")),
                                    list( "stoch_deter"          = as.character("Deterministic")),
                                    list( "theoretical_model"    = as.logical(FALSE)),
                                    list( "interventions_type"   = as.character("Vaccination")),
                                    list( "code_available"       = as.logical(TRUE)),
                                    list( "transmission_route"   = as.character("Sexual")),
                                    list( "assumptions"          = as.character("Unspecified")),
                                    list( "covidence_id"         = as.integer(2059))),
                                        vignette_prepend = "")

```

If you would like to add multiple models it is important to add each model first to the dataset before adding an additional one in order to ensure that the `model_data_id` is set correctly.

```{r eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
new_model_dataset <- append_new_entry_to_table(pathogen = 'marburg', table_type = 'model', new_row = new_model,
                          validate = TRUE, write_table = TRUE, vignette_prepend = "" )
```

## Step 4 -- add outbreak details

If the paper does not include a model you can skip this section. For this table we have a set number of field options which can be accepted. These field options can be viewed in the `Database field options` vignette. The validations in the function will assert that valid options have been used.

```{r add outbreak, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# load packages and source code
library(tidyverse)
library(validate)
source('../R/create_new_outbreak_entry.R')

new_outbreak <- create_new_outbreak_entry('marburg',new_outbreak = c( list( "article_id"           = as.integer(1)),
                                          list( "outbreak_start_day"   = as.integer(NA)),
                                          list( "outbreak_start_month" = as.character(NA)),
                                          list( "outbreak_start_year"  = as.integer(1999)),
                                          list( "outbreak_end_day"     = as.integer(NA)),
                                          list( "outbreak_end_month"   = as.character(NA)),
                                          list( "outbreak_date_year"   = as.integer(2001)),
                                          list( "outbreak_duration_months" = as.integer(NA)),
                                          list( "outbreak_size"        = as.integer(2)),
                                          list( "asymptomatic_transmission" = as.integer(0)),
                                          list( "outbreak_country"     = as.character("Tanzania")),
                                          list( "outbreak_location"    = as.character(NA)),
                                          list( "cases_confirmed"      = as.integer(NA)),
                                          list( "cases_mode_detection" = as.character(NA)),
                                          list( "cases_suspected"      = as.integer(NA)),
                                          list( "cases_asymptomatic"   = as.integer(NA)),
                                          list( "deaths"               = as.integer(2)),
                                          list( "cases_severe_hospitalised" = as.integer(NA)),
                                          list( "covidence_id"         = as.integer(2059)) ),
                                vignette_prepend = "")

```

If you would like to add multiple outbreaks it is important to add each outbreak first to the dataset before adding an additional one in order to ensure that the `outbreak_id` is set correctly.

```{r eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
new_outbreak_dataset <- append_new_entry_to_table(pathogen = 'marburg', table_type = 'outbreak', new_row = new_outbreak,
                          validate = TRUE, write_table = TRUE, vignette_prepend = "" )
```


## Step 5 -- add parameter details

If the paper does not include a model you can skip this section.

```{r add parameter, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# load packages and source code
library(tidyverse)
library(validate)
source('../R/create_new_parameter_entry.R')

new_parameter <- create_new_article_entry(pathogen = 'marburg',
                                     new_param   = c(  list( "article_id"            = as.integer(1)),
                                                       list( "parameter_type"        = as.character(NA)),
                                                       list( "parameter_value"       = as.double(NA)),
                                                       list( "parameter_unit"        = as.character(NA)),
                                                       list( "parameter_lower_bound" = as.double(NA)),
                                                       list( "parameter_upper_bound" = as.double(NA)),
                                                       list( "parameter_value_type"  = as.character(NA)),
                                                       list( "parameter_uncertainty_single_value" = as.double(NA)),
                                                       list( "parameter_uncertainty_singe_type"   = as.character(NA)),
                                                       list( "parameter_uncertainty_lower_value"  = as.double(NA)),
                                                       list( "parameter_uncertainty_upper_value"  = as.double(NA)),
                                                       list( "parameter_uncertainty_type"         = as.character(NA)),
                                                       list( "cfr_ifr_numerator"       = as.integer(NA)),
                                                       list( "cfr_ifr_denominator"     = as.integer(NA)),
                                                       list( "distribution_type"       = as.character(NA)),
                                                       list( "distribution_par1_value" = as.double(NA)),
                                                       list( "distribution_par1_type"  = as.character(NA)),
                                                       list( "distribution_par1_uncertainty" = as.logical(NA)),
                                                       list( "distribution_par2_value" = as.double(NA)),
                                                       list( "distribution_par2_type"  = as.character(NA)),
                                                       list( "distribution_par2_uncertainty" = as.logical(NA)),
                                                       list( "method_from_supplement"  = as.logical(NA)),
                                                       list( "method_moment_value"     = as.character(NA)),
                                                       list( "cfr_ifr_method"          = as.character(NA)),
                                                       list( "method_r"                = as.character(NA)),
                                                       list( "method_disaggregated_by" = as.character(NA)),
                                                       list( "method_disaggregated"    = as.logical(NA)),
                                                       list( "method_disaggregated_only" = as.logical(NA)),
                                                       list( "riskfactor_outcome"      = as.character(NA)),
                                                       list( "riskfactor_name"         = as.character(NA)),
                                                       list( "riskfactor_occupation"   = as.character(NA)),
                                                       list( "riskfactor_significant"  = as.character(NA)),
                                                       list( "riskfactor_adjusted"     = as.character(NA)),
                                                       list( "population_sex"          = as.character(NA)),
                                                       list( "population_sample_type"  = as.character(NA)),
                                                       list( "population_group"        = as.character(NA)),
                                                       list( "population_age_min"      = as.integer(NA)),
                                                       list( "population_age_max"      = as.integer(NA)),
                                                       list( "population_sample_size"  = as.integer(NA)),
                                                       list( "population_country"      = as.character(NA)),
                                                       list( "population_location"     = as.character(NA)),
                                                       list( "population_study_start_day"   = as.integer(NA)),
                                                       list( "population_study_start_month" = as.character(NA)),
                                                       list( "population_study_start_year"  = as.integer(NA)),
                                                       list( "population_study_end_day"     = as.integer(NA)),
                                                       list( "population_study_end_month"   = as.character(NA)),
                                                       list( "population_study_end_year"    = as.integer(NA)),
                                                       list( "genome_site"                  = as.character(NA)),
                                                       list( "genomic_sequence_available"   = as.logical(NA)),
                                                       list( "parameter_class"          = as.character(NA)),
                                                       list( "covidence_id"             = as.integer(2059)),
                                                       list( "Uncertainty"              = as.character(NA)),
                                                       list( "Survey year"              = as.character(NA))),
                                      vignette_prepend = "")

```

If you would like to add multiple parameters it is important to add each parameter first to the dataset before adding an additional one in order to ensure that the `parameter_data_id` is set correctly.

```{r eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
new_parameter_dataset <- append_new_entry_to_table(pathogen = 'marburg', table_type = 'parameter', new_row = new_parameter,
                          validate = TRUE, write_table = TRUE, vignette_prepend = "" )
```


## Step 6 -- create a pull request

Once you have completed the above steps to add all required data for a new paper you can commit the updates to the dataset on your branch and push your branch. Please submit a pull request with the following

- [ ] Details of the paper, including doi, and that you have confirmed that this paper is in line with inclusion and exclusion criteria of the systematic review
- [ ] What pathogen this update concerns and what tables you have updated
- [ ] Any comments / observations the reviewer may find helpful

We strongly encourage the use of the utility functions set out in this vignette. We will run the data again through these validation tests and if they fail the pull request will be rejected.
